---
title: "Project 2: Walmart Store Sales Forcasting"
date: "Fall 2020"
output:
  html_notebook:
    theme: readable
    toc: TRUE
    toc_float: TRUE
---

Historical sales data for 45 Walmart stores located in different regions is provided. Each store contains many departments. The goal is to predict the future weekly sales for each department in each store based on the historical data.

## Source

You can find relevant information and useful discussion and sample code on Kaggle [https://www.kaggle.com/c/walmart-recruiting-store-sales-forecasting](https://www.kaggle.com/c/walmart-recruiting-store-sales-forecasting). Note that **ONLY** the training data is used in this project and our evaluation procedure is **different** from the one on Kaggle.


## Datasets

Download the zip file, `train.csv.zip`. Then use the following code to generate datasets  for this project. 

```{r eval=, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)

# read raw data and extract date column
train_raw <- readr::read_csv(unz('train.csv.zip', 'train.csv'))
train_dates <- train_raw$Date

# training data from 2010-02 to 2011-02
start_date <- ymd("2010-02-01")
end_date <- start_date %m+% months(13)

# split dataset into training / testing
train_ids <- which(train_dates >= start_date & train_dates < end_date)
train = train_raw[train_ids, ]
test = train_raw[-train_ids, ]

# create the initial training data
readr::write_csv(train, 'train_ini.csv')

# create test.csv 
# removes weekly sales
test %>% 
  select(-Weekly_Sales) %>% 
  readr::write_csv('test.csv')

# create 10-fold time-series CV
num_folds <- 10
test_dates <- train_dates[-train_ids]

# month 1 --> 2011-03, and month 20 --> 2012-10.
# Fold 1 : month 1 & month 2, Fold 2 : month 3 & month 4 ...
for (i in 1:num_folds) {
    # filter fold for dates
    start_date <- ymd("2011-03-01") %m+% months(2 * (i - 1))
    end_date <- ymd("2011-05-01") %m+% months(2 * (i - 1))
    test_fold <- test %>%
        filter(Date >= start_date & Date < end_date)
    
    # write fold to a file
    readr::write_csv(test_fold, paste0('fold_', i, '.csv'))
}
```

The code above will generate the following files:

* **train_ini.csv**: 5 columns ("Store", "Dept", "Date", "Weekly_Sales", "IsHoliday"), same as the train.csv file on Kaggle but ranging from 2010-02 to 2011-02.

* **test.csv**: 4 columns ("Store", "Dept", "Date", "IsHoliday"), in the same format as the train.csv file on Kaggle ranging from 2011-03 to 2012-10 with the "Weekly_Sales" column being removed.

* **fold_1.csv**, ..., **fold_10.csv**: 5 columns ("Store", "Dept", "Date", "Weekly_Sales", "IsHoliday"), same as the train.csv file on Kaggle, and one for every two months starting from 2011-03 to 2012-10.


## Goal

The file, train_ini.csv, provides the weekly sales data for various stores and departments from 2010-02 (February 2010) to 2011-02 (February 2011).

Given train_ini.csv, the data till 2011-02, need to predict the weekly sales for 2011-03 and 2011-04. Then you'll be provided with the weekly sales data for 2011-03 and 2011-04 (fold_1.csv), and you need to predict the weekly sales for 2011-05 and 2011-06, and so on:

* `t = 1`, predict 2011-03 to 2011-04 based on data from 2010-02 to 2011-02 (train_ini.csv); 
* `t = 2`, predict 2011-05 to 2011-06 based on data from 2010-02 to 2011-04 (train_ini.csv, fold_1.csv);
* `t = 3`, predict 2011-07 to 2011-08 based on data from 2010-02 to 2011-06 (train_ini.csv, fold_1.csv, fold_2.csv);
* ......
* `t = 10`, predict 2011-09 to 2011-08 baesd on data from 2010-02 to 2011-08 (train_ini.csv, fold_1.csv, fold_2.csv, ..., fold_9.csv)

## Models

```{r}
naive = function(){
  if (t>1){
    train <<- train %>% add_row(new_train)
  }
  
  start_date <- ymd("2011-03-01") %m+% months(2 * (t - 1))
  end_date <- ymd("2011-05-01") %m+% months(2 * (t - 1))
  test_current <- test %>%
    filter(Date >= start_date & Date < end_date) %>%
    select(-IsHoliday)
  
  most_recent_date <- max(train$Date)
  tmp_train <- train %>%
    filter(Date == most_recent_date) %>%
    rename(Weekly_Pred = Weekly_Sales) %>%
    select(-Date, -IsHoliday)
  
  test_pred <- test_current %>%
    left_join(tmp_train, by = c('Dept', 'Store')) 
  return(test_pred)
}
```

```{r}
linearmodelmethod = function(){
  
  start_date <- ymd("2011-03-01") %m+% months(2 * (t - 1))
  end_date <- ymd("2011-05-01") %m+% months(2 * (t - 1))
  test_current <- test %>%
    filter(Date >= start_date & Date < end_date) %>%
    select(-IsHoliday)
  
  if (t>1){
    train <<- train %>% add_row(new_train)
  }
  # not all depts need prediction
  test_depts <- unique(test_current$Dept)
  test_pred <- NULL
  
  for(dept in test_depts){
    train_dept_data <- train %>% filter(Dept == dept)
    test_dept_data <- test_current %>% filter(Dept == dept)
    
    # no need to consider stores that do not need prediction
    # or do not have training samples
    train_stores <- unique(train_dept_data$Store)
    test_stores <- unique(test_dept_data$Store)
    test_stores <- intersect(train_stores, test_stores)
    
    for(store in test_stores){
      tmp_train <- train_dept_data %>% 
        filter(Store == store) %>%
        mutate(Wk = ifelse(year(Date) == 2010, week(Date)-1, week(Date))) %>%
        mutate(Yr = year(Date))
      tmp_test <- test_dept_data %>% 
        filter(Store == store) %>%
        mutate(Wk = ifelse(year(Date) == 2010, week(Date)-1, week(Date))) %>%
        mutate(Yr = year(Date)) 
      
      tmp_train$Wk = factor(tmp_train$Wk, levels = 1:52)
      tmp_test$Wk = factor(tmp_test$Wk, levels = 1:52)
      
      train_model_matrix <- model.matrix(~ Yr + Wk, tmp_train)
      test_model_matrix <- model.matrix(~ Yr + Wk, tmp_test)
      mycoef <- lm(tmp_train$Weekly_Sales ~ train_model_matrix)$coef
      mycoef[is.na(mycoef)] <- 0
      tmp_pred <- mycoef[1] + test_model_matrix %*% mycoef[-1]
      
      tmp_test_wk <- tmp_test
      test_pred_wk <- test_pred
      
      tmp_test <- tmp_test %>%
        mutate(Weekly_Pred = tmp_pred[,1]) %>%
        #select(-Yr)
        select(-Yr)
      test_pred <- test_pred %>% bind_rows(tmp_test)  
      
    }
  }
  #do a circular shift for fold5
  if (t==5){
    test_pred<-shift(test_pred_wk)
  }
  
  #test_pred<-test_pred %>% 
  # select(-Wk)
  return(test_pred)
}
```

```{r}
naive2 = function(){
  
  start_date <- ymd("2011-03-01") %m+% months(2 * (t - 1))
  end_date <- ymd("2011-05-01") %m+% months(2 * (t - 1))
  test_current <- test %>%
    filter(Date >= start_date & Date < end_date) %>%
    select(-IsHoliday)
  
  if (t>1){
    train <<- train %>% add_row(new_train)
  }
  
  start_last_year = min(test_current$Date) - 375
  end_last_year = max(test_current$Date) - 350
  tmp_train <- train %>%
    filter(Date > start_last_year & Date < end_last_year) %>%
    mutate(Wk = ifelse(year(Date) == 2010, week(Date)-1, week(Date))) %>%
    rename(Weekly_Pred = Weekly_Sales) %>%
    select(-Date, -IsHoliday)
  
  test_current <- test_current %>%
    mutate(Wk = week(Date))
  
  test_pred <- test_current %>%
    left_join(tmp_train, by = c('Dept', 'Store', 'Wk')) %>%
    select(-Wk)
  
  test_pred <- test_current %>%
    left_join(tmp_train, by = c('Dept', 'Store', 'Wk')) 
  #  select(-Wk)
  ###Add shift
  if(t==5){
    test_pred_wk<-test_current %>%
      left_join(tmp_train, by = c('Dept', 'Store', 'Wk'))
    
    test_pred<-shift(test_pred_wk)
  }
  return(test_pred)
}
```

```{r}
shift=function(test_pred_wk){
  
  weeks2shift=c(49,50,51,52)
  days2shift=1
  test_pred_shift<-test_pred_wk
  for (indx in 1:length(weeks2shift)){
    prevwk <- test_pred_shift %>%
      filter(Wk==(weeks2shift[indx]-1)) %>%
      mutate(offset1=round(Weekly_Pred*days2shift/7,2)) %>%
      select(-Weekly_Pred, -Date, -Wk)
    
    
    nextwk <- test_pred_shift %>%
      filter(Wk==(weeks2shift[indx])) %>%
      mutate(offset2=round(Weekly_Pred*(7-days2shift)/7,2)) %>%
      select(-Weekly_Pred)
    
    combine<-prevwk %>%
      right_join(nextwk, by = c('Store', 'Dept')) %>%
      mutate(Weekly_Pred=offset1+offset2) %>%
      select(-offset1,-offset2)
    
    test_pred_shift<-test_pred_shift%>%
      filter(Wk!=(weeks2shift[indx]))
    test_pred_shift <- rbind(test_pred_shift,combine)
    
  }  
  
  return(test_pred_shift)
}
```

```{r}
mypredict=function(){
  linear_pred=linearmodelmethod()
  linear_pred<-linear_pred %>% select(-Wk)
  snaive_pred=naive2()
  snaive_pred<-snaive_pred%>% select(-Wk)
  if((t==4)||(t==5)||(t==6)||(t==10)){
    test_pred<-linear_pred %>%
      right_join(snaive_pred, by = c('Store', 'Dept','Date')) %>%
      mutate(Weekly_Pred=((Weekly_Pred.x+Weekly_Pred.y)/2)) 
  }
  else
    test_pred<-linear_pred
  return(test_pred)
}
```

## Code Evaluation

```{r message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)


#source("mymain.R")

# read in train / test dataframes
train <- readr::read_csv('train_ini.csv')
test <- readr::read_csv('test.csv')

# save weighted mean absolute error WMAE
num_folds <- 10
wae <- rep(0, num_folds)

for (t in 1:num_folds) {
  # *** THIS IS YOUR PREDICTION FUNCTION ***
  test_pred <- mypredict()
  
  # load fold file 
  fold_file <- paste0('fold_', t, '.csv')
  new_train <- readr::read_csv(fold_file, 
                               col_types = cols())

  # extract predictions matching up to the current fold
  scoring_tbl <- new_train %>% 
      left_join(test_pred, by = c('Date', 'Store', 'Dept'))
  
  # compute WMAE
  actuals <- scoring_tbl$Weekly_Sales
  preds <- scoring_tbl$Weekly_Pred
  preds[is.na(preds)] <- 0
  weights <- if_else(scoring_tbl$IsHoliday, 5, 1)
  wae[t] <- sum(weights * abs(actuals - preds)) / sum(weights)
}

print(wae)
mean(wae)
```


We use the same evaluation metric as the [one described on Kaggle](https://www.kaggle.com/c/walmart-recruiting-store-sales-forecasting/overview/evaluation), with higher weihts on holiday weeks. There are four holiday weeks per year:

* Super Bowl
* Labor Day
* Thanksgiving
* Christmas





